Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-07-02 submission/methods.tex
full pass in light of kdd reviewer comments

Let us write the Energy Landscape (or optimization function, parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, activation functions $h_{l}(\cdot)$, and $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$,~as:
\begin{equation}
E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})  .
\label{eqn:dnn_energy}
\end{equation}
Each DNN layer contains one or more layer 2D  $N\times M$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, extracted from 2D Convolutional layers, and where $N > M$.% 
\footnote{We do not use intra-layer information from the models in our quality metrics, but (as we will describe) our metrics can be used to learn about intra-layer model properties.}
(We may drop the $i$ and/or $i,l$ subscripts below.)
See Appendix~\ref{sxn:appendix} for how we define the Conv2D layer matrixes and for our choices of normalization.

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretrained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.

@@

The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.%

@@

\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum_{i=1}^{M} \lambda_{i}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$
\item 
Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm):%
\footnote{Notice $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$. We use $\mathbf{X}$ to emphasize that $\alpha$ depends on the ESD of $\mathbf{X}$.}
 $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
Recall that the eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$: $\lambda_{i}=\sigma^{2}_{i}$.
Also, note that we do \emph{not} normalize $\mathbf{X}$ by $1/N$; see Appendix~\ref{sxn:appendix} for a discussion of this issue.

The first two norms are well-known in ML; the last two deserve special mention.
The empirical parameter $\alpha$ is the Power Law (PL) exponent that arises in the recently-developed HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Operationally, $\alpha$ is determined by using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}.
\end{equation}
%where $\lambda_{max}$ \emph{is} the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
%(and it approximates the average log $\alpha$-Shatten Norm metric),
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a limited set of models.

@@

\noindent
For small $\alpha$, the Weighted Alpha metric approximates the Log $\alpha$-Shatten norm, as can be shown with a statistical mechanics and random matrix theory derivation \cite{MM20_unpub_work}; and the Weighted Alpha and $\alpha$-Shatten norm metrics often behave like an improved, weighted average Log Spectral Norm, and may track this metric in some~cases.

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.  
(Technically, we also need the minimum eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.)
The PL exponent $\alpha$ measures of the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
%
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric for the entire~DNN.

%KDD% \vspace{-1mm}
\paragraph{Convolutional Layers and Normalization issues.}
There are several technical issues (regarding spectral analysis of convolutional layers and normalization of empirical matrices) that are important for reproducibility of our results.
See Appendix~\ref{sxn:appendix} for a discussion.

>>>>

Let us write the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, activation functions $h_{l}(\cdot)$, and $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$,~as:
\begin{equation}
E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})  .
\label{eqn:dnn_energy}
\end{equation}
Each DNN layer contains one or more layer 2D  $N\times M$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, extracted from 2D Convolutional layers, and where $N > M$.% 
\footnote{We do not use intra-layer information from the models in our quality metrics, but (as we will describe) our metrics can be used to learn about intra-layer model properties.}
(We may drop the $i$ and/or $i,l$ subscripts below.)
See Appendix~\ref{sxn:appendix} for how we define the Conv2D layer matrixes and for our choices of normalization.

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretrained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.
Our methodological approach is thus similar to a statistical meta-analysis, common in biomedical research.

@@

Since we analyze pretrained models, most of our analysis involves analyzing DNN weight matrices.
The best performing empirical quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's \emph{Empirical Correlation Matrix}: $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.%

@@

\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum_{i=1}^{M} \lambda_{i}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$
\item 
Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm):%
\footnote{Notice $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}$. We use $\mathbf{X}$ to emphasize that $\alpha$ depends on the ESD of $\mathbf{X}$.}
 $\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
Recall that the eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$: $\lambda_{i}=\sigma^{2}_{i}$.
Also, note that we do \emph{not} normalize $\mathbf{X}$ by $1/N$.
See Appendix~\ref{sxn:appendix} for a discussion of this issue.
Observe that all four metrics can be computed easily from DNN weight matrices.

The first two metrics are well-known in ML; the last two deserve special mention.
The empirical parameter $\alpha$ is the Power Law (PL) exponent that arises in the recently-developed HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
The bottom line from HT-SR Theory is that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models; see Appendix~\ref{sxn:theory-review-appendix} for more details.
Operationally, $\alpha$ is determined by analyzing DNN weight matrices using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}.
\end{equation}
%where $\lambda_{max}$ \emph{is} the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
%(and it approximates the average log $\alpha$-Shatten Norm metric),
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced and justified previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a much smaller and more limited set of models.

@@

\noindent
For small $\alpha$, the Weighted Alpha metric approximates the Log $\alpha$-Shatten norm, as can be shown with a statistical mechanics and random matrix theory derivation \cite{MM20_unpub_work}; and the Weighted Alpha and $\alpha$-Shatten norm metrics often behave like an improved, weighted average Log Spectral Norm.

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.% 
\footnote{Technically, we also need the minimum eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.}
The PL exponent $\alpha$ measures the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
%
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric for the entire~DNN.

%KDD% \vspace{-1mm}
\paragraph{Convolutional Layers and Normalization issues.}
There are several technical issues, e.g., regarding spectral analysis of convolutional layers and normalization of empirical matrices, that are important for reproducibility of our results.
See Appendix~\ref{sxn:appendix} for a discussion.

### 2020-07-02 submission/nlp_models.tex
full pass in light of kdd reviewer comments

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
%, both in terms of their architecture (e.g., whether there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our empirical quality metrics.

While norm-based metrics perform reasonably well on well-trained NLP models, they often behave anomalously on poorly-trained models.
Indeed, for such ``bad'' models, weight matrices may display rank collapse, decreased Frobenius mass, or unusually small Spectral norms.
(This may be misinterpreted as ``smaller is better.'')
In contrast, PL-based metrics, including the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $) display consistent behavior, even on poorly trained models.
Indeed, we can use these metrics to help identify when architectures need repair and when more and/or better data are needed.

@@

Many NLP models, such as GPT and BERT, have some weight matrices with unusually large PL exponents (e.g., $\alpha\gg 6$).
This indicates these matrices may be \emph{under}-correlated (i.e., over-parameterized, relative to the amount of data).
In this regime, the truncated PL fit itself may not be very reliable because the MLE estimator it uses is unreliable in this range (i.e., the specific $\alpha$ values returned by the truncated PL fits are less reliable, but having large versus small values of $\alpha$ is reliable).
Phenomenologically, if we examine the ESD visually, we can usually describe these $\mathbf{W}$ as in the \emph{Bulk-Decay} or \emph{Bulk-plus-Spikes} phase~\cite{MM18_TR,MM19_HTSR_ICML}.
Previous work~\cite{MM18_TR,MM19_HTSR_ICML} has conjectured that very well-trained DNNs would not have many \emph{outlier} $\alpha\gg 6$; and improved versions of GPT (shown below) and BERT (not shown) confirm this.

@@

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can identify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

@@

We have analyzed the four quality metrics described in Section~\ref{sxn:methods} for the OpenAI GPT and GPT2 pretrained models.
See Table \ref{table:nlp} for a summary of results.
We start by examining trends between GPT and GPT2-small.
Observe that all four metrics increase when going from GPT to GPT2-small, i.e., they are smaller for the higher-quality model (higher quality since GPT was trained to better data), when the number of layers is held fixed.
Notice that in the GPT model, being poorly trained, the norm metrics all exhibit \emph{Scale Collapse}, compared to GPT2-small.

We next examine trends between GPT2-medium to GPT2-large to GPT2-xl.
Observe that (with one minor exception involving the log Frobenius norm metric) all four metrics decrease as one goes from medium to large to xl, indicating that the larger models indeed look better than the smaller models.
Notice that, for these well-trained models, the norm metrics now behave as expected, decreasing with increasing accuracy.

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unusually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We next describe the behavior of the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the ``bad'' GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the ``good'' GPT2-small, violating the conventional wisdom that smaller Spectral norms are better.
Indeed, because there are so many anonymously small Spectral norms, it appears that the GPT model may be exhibiting a kind of \emph{Scale Collapse}, like that observed in the distilled CV models (in Figure~\ref{fig:resnet204D5L}).
This is important because it demonstrates that, while the Spectral (or Frobenius) norm may correlate well with predicted test error, it is \emph{not} a good indicator of the overall model quality.
It can mispredict good-versus-bad questions in ways not seen with PL-based metrics.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient models.

@@

(Note that Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
%That is, they correspond to the change in normalization of the word embedding layer(s), and thus they can be ignored.
These layers have a different effective normalization, and therefore a different scale.
% \charles{We discuss this further in Appendix~\ref{sxn:appendix}.}
% For example, in GPT, for most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$, but in the first embedding layer, the maximum is of order $N$ (the number of words in the embedding), or $\lambda_{max}\sim\mathcal{O}(10^{5})$.  
% For GPT and GPT2, we treat all layers as-is (although one may want to normalize the first 2 layers by $\mathbf{X}$ by $1/N$ or treat them as an outlier).
We discuss this further in Appendix~\ref{sxn:appendix}.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.)

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistent and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models.
These results show that the Correlation Flow differs significantly between GPT and GPT2-small (with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}).

These results should be contrasted with the corresponding results for Spectral Norms, shown in Figure~\ref{fig:gpt-snorm-layer}.
Attention models have two types of layers, one small and large; and the Spectral Norm, in particular, displays unusually small values for some of these layers for GPT.
This Scale Collapse for the poorly-trained GPT is similar to what we observed for the distilled ResNet20 model in Figure~\ref{fig:resnet204Dalpha}.
Because of the anomalous scale collapse that is frequently observed in poorly-trained models, these results suggest that scale-dependent norm metrics should not be directly applied to distinguish good-versus-bad models.

@@

We now look across series of increasingly improving GPT2 models (i.e., we consider good-better-best questions), by examining both the PL exponent $\alpha$ as well as the Log Norm metrics.  
In general, as we move from GPT2-medium to GPT2-xl, histograms for both $\alpha$ exponents and the Log Norm metrics downshift from larger to smaller values. 
For example, see Figure \ref{fig:gpt2-histograms}, which shows the histograms over the layer weight matrices for fitted PL exponent ($\alpha$) and the Log Alpha Norm ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$)~metric.

We see that the average $\alpha$ decreases with increasing model size, although the differences are less noticeable between the differing good-better-best GTP2 models than between the good-versus-bad GPT and GPT2-small models.
Unlike GPT, however, the layer Log Alpha Norms behave more as expected for GPT2 layers, with the larger models consistently having smaller norms. 
Similarly, the Log Spectral Norm also decreases on average with the larger models (not shown).  
As expected, the norm metrics can indeed distinguish among good-better-best models among a series well-trained models.

>>>>

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the information correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
%, both in terms of their architecture (e.g., whether there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our empirical quality metrics.

While norm-based metrics perform reasonably well on well-trained NLP models, they often behave anomalously on poorly-trained models.
Indeed, for such ``bad'' models, weight matrices may display rank collapse, decreased Frobenius mass, or unusually small Spectral norms.
This may be misinterpreted as ``smaller is better.'' 
Instead, it should probably be interpreted as being due to a similar mechanism to how distillation can ``damage'' otherwise good models.
In contrast to norm-based metrics, PL-based metrics, including the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $) display consistent behavior, even on poorly trained models.
Indeed, we can use these metrics to help identify when architectures need repair and when more and/or better data are needed.

@@

We have observed that many NLP models, such as GPT and BERT, have some weight matrices with unusually large PL exponents (e.g., $\alpha\gg 6$).
This indicates these matrices may be \emph{under}-correlated (i.e., over-parameterized, relative to the amount of data).
In this regime, the truncated PL fit itself may not be very reliable because the MLE estimator it uses is unreliable in this range.
That is, the specific $\alpha$ values returned by the truncated PL fits are less reliable, but having large versus small values of $\alpha$ is reliable.
Phenomenologically, if we examine the ESD visually, we can usually describe these $\mathbf{W}$ as in the \emph{Bulk-Decay} or \emph{Bulk-plus-Spikes} phase from HT-ST Theory~\cite{MM18_TR,MM19_HTSR_ICML}.
Previous work~\cite{MM18_TR,MM19_HTSR_ICML} has conjectured that very well-trained DNNs would not have many \emph{outlier} $\alpha\gg 6$; and, consistent with this, improved versions of GPT (shown below) and BERT (not shown) confirm this.

@@

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can identify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously, as well as analyze well-trained versus more-well-trained models.

@@

We have analyzed the four quality metrics described in Section~\ref{sxn:methods} for the OpenAI GPT and GPT2 pretrained models.
See Table \ref{table:nlp} for a summary of results.
We start by examining trends between GPT and GPT2-small.
Observe that all four metrics increase when going from GPT to GPT2-small, i.e., they are smaller for the higher-quality model (higher quality since GPT was trained to better data), when the number of layers is held fixed.
Notice that in the poorly-trained GPT model, the norm metrics all exhibit \emph{Scale Collapse}, compared to well-trained GPT2-small model.

We next examine trends between GPT2-medium to GPT2-large to GPT2-xl.
Observe that (with one minor exception involving the log Frobenius norm metric) all four metrics decrease as one goes from medium to large to xl, indicating that the larger models indeed look better than the smaller models.
Notice that, for these well-trained models, the norm metrics now behave as expected, i.e., there is no \emph{Scale Collapse} and norms are decreasing with increasing~accuracy.

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unusually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the poorly-trained GPT versus the well-trained GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We next describe the behavior of the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the poorly-trained GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the well-trained GPT2-small, violating the conventional wisdom that smaller Spectral norms are better.
Indeed, because there are so many anonymously small Spectral norms, it appears that the GPT model may be exhibiting a kind of \emph{Scale Collapse}, like that observed in the distilled CV models (in Figure~\ref{fig:resnet204D5L}).

This is important because it demonstrates that, while the Spectral (or Frobenius) norm may correlate well with predicted test error, at least among reasonably well-trained models, it is \emph{not} a good indicator of the overall model quality in general.
In particular, it can mispredict well-trained versus poorly-trained questions in ways not seen with PL-based metrics.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient models.

@@

We should note that Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
%That is, they correspond to the change in normalization of the word embedding layer(s), and thus they can be ignored.
These layers have a different effective normalization, and therefore a different scale.
% \charles{We discuss this further in Appendix~\ref{sxn:appendix}.}
% For example, in GPT, for most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$, but in the first embedding layer, the maximum is of order $N$ (the number of words in the embedding), or $\lambda_{max}\sim\mathcal{O}(10^{5})$.  
% For GPT and GPT2, we treat all layers as-is (although one may want to normalize the first 2 layers by $\mathbf{X}$ by $1/N$ or treat them as an outlier).
We discuss this further in Appendix~\ref{sxn:appendix}.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistent and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models.
These results suggest that the Correlation Flow is also a useful concept for NLP models, and that it differs significantly between GPT and GPT2-small, with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}.

These results should be contrasted with the corresponding results for Spectral Norms, shown in Figure~\ref{fig:gpt-snorm-layer}.
Attention models have two types of layers, one small and large; and the Spectral Norm, in particular, displays unusually small values for some of these layers for GPT.
This Scale Collapse for the poorly-trained GPT is similar to what we observed for the distilled ResNet20 model in Figure~\ref{fig:resnet204Dalpha}.
Because of the anomalous scale collapse that is frequently observed in poorly-trained models, these results suggest that scale-dependent norm metrics should not be directly applied to distinguish well-trained versus poorly-trained models.

@@

We now look across series of increasingly improving GPT2 models (i.e., we consider well-trained versus more-well-trained questions), by examining both the PL exponent $\alpha$ as well as the Log Norm metrics.  
In general, as we move from GPT2-medium to GPT2-xl, histograms for both $\alpha$ exponents and the Log Norm metrics downshift from larger to smaller values. 
For example, see Figure \ref{fig:gpt2-histograms}, which shows the histograms over the layer weight matrices for fitted PL exponent ($\alpha$) and the Log Alpha Norm ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$)~metric.

We see that the average $\alpha$ decreases with increasing model size, although the differences are less noticeable between the differing well-trained versus more-well-trained GTP2 models than between the poorly-trained versus well-trained GPT and GPT2-small models.
Unlike GPT, however, the layer Log Alpha Norms behave more as expected for GPT2 layers, with the larger models consistently having smaller norms. 
Similarly, the Log Spectral Norm also decreases on average with the larger models (not shown).  
As expected, the norm metrics can indeed distinguish among well-trained versus more-well-trained models among a series well-trained models.
